Lossy compression of DNA sequencing quality data

### 2015-02-03 manuscript/q-compression.tex
Revised text, including removal of text regarding image compression.

The last approach we investigated is based on transforming the
original data into a wholly different representation for which
efficient and proven compression algorithms already exist. In this
study, we described a way of reducing the problem of quality
compression to image compression via block-wise quality data
transformation. This approach is different from the others in that
depending on the image compression algorithm used it is not restricted
to lossy techniques.

@@

We compare the \textsc{mse} versus \emph{bits per bp} of the fragment
and short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
data set (Figures \ref{fig:mse_vs_bpbp_frag} and
\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
requires 1 \emph{byte} per basepair because they are stored in
\textsc{ascii} format; however, losslessly compressing the values with
BZip2 2.112 bits per basepair. 0-order polynomial regression and
32-profile encodings have the lowest bits per bp. 5th-order polynomial
regression and \textsc{png} encodings have the highest bits per bp,
but have among the lowest \textsc{mse}. Despite having similar bits
per bp, \textsc{png} encoding has roughly one third the \textsc{mse}
as 5th-order polynomial (3.07 vs 10.27, respectively).

As the order of the polynomial increases, the bits per bp increase and
the \textsc{mse} decreases at an exponential rate.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{MSE versus bits per bp for different compression methods
  applied to the fragment library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. PNG:
  Blockwise image compres the generated SAM file for each compressing
  approach weresion algorithm. Dashed line corresponds to lossless
  bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{MSE versus bits per bp for for different compression methods
  applied to the short-jump library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. PNG:
  Blockwise image compression algorithm. Dashed line corresponds to
  lossless bzip compression of quality
  values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

@@

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using good/bad encoding, 0--5th-order polynomial regression encodings,
32- and 256-profile encodings, and \textsc{png} encoding. We evaluate
the assembly based on contiguity statistics, log average read
probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
LAP score can be viewed as a log likelihood score, where a value
closer to 0 is better. We use a reference-based evaluation script
provided by \textsc{gage} which counts single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to aid in comparison. Alongside the raw metrics, we also provide a
cummulative sum of the individual ranks of each method for each
metric. A similar approach was utilized by the \textsc{Assemblathon1}
competition~\cite{Earl:2011fv}.

The assembly using the uncompressed reads outperforms all compression
methods in terms of LAP, NG50, relocations, translations, and
inversions (Table \ref{tab:assembly_ranks}). Among the compression
methods, the \textsc{png} encoding had the highest rank, followed by
256- and 32-profile encoding, then the 4th-order polynomial
regression, the 5th-order polynomial regression, the good/bad
encoding, and lastly, the 3rd through 0-order polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of SNPs. The order of polynomial has an
inverse relationship with the amount of SNPs detected. The good/bad,
profile and PNG methods detected the least amount of SNPs compared to
the reference genome, outperforming the assembly using the original
quality sequences. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual SNPs.

@@

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
PNG                    & 2    & 4    & 1          & 2           & 2              & 3      & 4          & 2                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

Generally, increasing the regression model polynomial order from 0 to
5 results in more mapped reads with the most marked change from order
1 to order 2. The best compression approach is the one that has
highest proportion of reads shared between the uncompressed and
decompressed files and least number of reads that are unique in both
uncompressed and decompressed files. In fragment sample, on average,
99.03\% of reads were mapped using different compression approaches,
within which 99\% are shared with the uncompressed alignments and
0.02\% unique mapped reads in the decompressed file. The average
missed alignments are 0.99\%. Compression with profile 256 approach
was shown to be the best performing approach in the fragment
decompressed sample, as a proportional of 99.94\% reads are mapped
with respect to the uncompressed mapped reads within which 99.91\% are
common and only 0.08\% unique mapped reads in the decompressed
file. In short-run sample, the average number of mapped reads is 99.14
\% within which 98.73\% are common and 0.41\% unique mapped reads in
the decompressed file.The average missed alignments are
1.2\%. Regression with the 5th polynomial order captured the highest
mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
unique mapped reads in the decompressed file. In both samples,
\textsc{png} compression has the least number of mapped reads that are
unique in the decompressed files (only 0.013\% and 0.89\% in the
fragments and short-jump samples, respectively).

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie 2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{Percentage of reads mapped}\\
    \cline{3-6}
	Library & Compression Strategy & Total & Shared & Uncompressed only & Compressed only \\ \hline
	& Maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\
	&Good/Bad & 99.92 & 99.89 & 0.11 & 0.03 \\
	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&Regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&Regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	Fragment&Regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&Regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&Regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&Profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
	&PNG & 99.82 & 99.81 & 0.19 & 0.01 \\

@@

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&Good/Bad & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
	&PNG & 99.71 & 99.62 & 0.38 & 0.09 \\ \hline
\end{tabular}

@@

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  proportion of bases/reads that are missed or gained after applying
  compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{Fragments} & \multicolumn{2}{c}{Short-Jump} \\
	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
	Good/bad & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	Regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	Regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	Regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	Regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	Regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	Profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
	PNG & 99.62 & 99.75 & 99.84 & 99 \\ \hline
\end{tabular}
\end{table}

>>>>

@@

We compare the \textsc{mse} versus \emph{bits per bp} of the fragment
and short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
data set (Figures \ref{fig:mse_vs_bpbp_frag} and
\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
requires 1 \emph{byte} per basepair because they are stored in
\textsc{ascii} format; however, lossless compression of the values with
BZip2 2.112 bits per basepair. 0-order polynomial regression and
32-profile encodings have the lowest bits per bp. 5th-order polynomial
regression has the highest bits per bp,
but have among the lowest \textsc{mse}. As the order of the polynomial increases, the bits per bp increase and
the \textsc{mse} decreases at an exponential rate.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{MSE versus bits per bp for different compression methods
  applied to the fragment library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to lossless
  bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{MSE versus bits per bp for for different compression methods
  applied to the short-jump library from \textit{Rhodobacter
    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
  Quality profile compression with 32 and 256 profiles. R0-R5:
  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to
  lossless bzip compression of quality
  values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

@@

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using good/bad encoding, 0--5th-order polynomial regression encodings,
and 32- and 256-profile encodings. We evaluate
the assembly based on contiguity statistics, log average read
probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
LAP score can be viewed as a log likelihood score, where a value
closer to 0 is better. We use a reference-based evaluation script
provided by \textsc{gage} which counts single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to aid in comparison. Alongside the raw metrics, we also provide a
cummulative sum of the individual ranks of each method for each
metric. A similar approach was utilized by the \textsc{Assemblathon1}
competition~\cite{Earl:2011fv}.

The assembly using the uncompressed reads outperforms all compression
methods in terms of LAP, NG50, relocations, translations, and
inversions (Table \ref{tab:assembly_ranks}). Among the compression
methods, the 
256- and 32-profile encoding had the highest rank, then the 4th-order polynomial
regression, the 5th-order polynomial regression, the good/bad
encoding, and lastly, the 3rd through 0-order polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of SNPs. The order of polynomial has an
inverse relationship with the amount of SNPs detected. The good/bad and
profile methods detected the least amount of SNPs compared to
the reference genome, outperforming the assembly using the original
quality sequences. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual SNPs.

@@

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

Generally, increasing the regression model polynomial order from 0 to
5 results in more mapped reads with the most marked change from order
1 to order 2. The best compression approach is the one that has
highest proportion of reads shared between the uncompressed and
decompressed files and least number of reads that are unique in both
uncompressed and decompressed files. In fragment sample, on average,
99.03\% of reads were mapped using different compression approaches,
within which 99\% are shared with the uncompressed alignments and
0.02\% unique mapped reads in the decompressed file. The average
missed alignments are 0.99\%. Compression with profile 256 approach
was shown to be the best performing approach in the fragment
decompressed sample, as a proportional of 99.94\% reads are mapped
with respect to the uncompressed mapped reads within which 99.91\% are
common and only 0.08\% unique mapped reads in the decompressed
file. In short-run sample, the average number of mapped reads is 99.14
\% within which 98.73\% are common and 0.41\% unique mapped reads in
the decompressed file.The average missed alignments are
1.2\%. Regression with the 5th polynomial order captured the highest
mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
unique mapped reads in the decompressed file.

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie 2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{Percentage of reads mapped}\\
    \cline{3-6}
	Library & Compression Strategy & Total & Shared & Uncompressed only & Compressed only \\ \hline
	& Maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\
	&Good/Bad & 99.92 & 99.89 & 0.11 & 0.03 \\
	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&Regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&Regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	Fragment&Regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&Regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&Regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&Profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\

@@

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&Good/Bad & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  proportion of bases/reads that are missed or gained after applying
  compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{Fragments} & \multicolumn{2}{c}{Short-Jump} \\
	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
	Good/bad & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	Regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	Regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	Regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	Regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	Regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	Profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
\end{tabular}
\end{table}
